---
title: "Black Friday Dataset from AV"
author: "KISHORE"
date: "January 28, 2017"
output: html_document
---
### Data explored here istaken from Analytics Vidya practice problem.
### The code to create new features is taken from another member (vopani)

loading libraries
```{r}
library(dummies)
library(plyr)
library(ggplot2)
```


```{r}
train <- read.csv("C:/Users/Admin/Downloads/Black Friday/train_oSwQCTC/train.csv", stringsAsFactors=F)
```


onehot-encoding city variable
```{r}
X_train <- train
X_train <- dummy.data.frame(X_train, names=c("City_Category"), sep="_")
```


Converting age variable to numeric
```{r}
X_train$Age[X_train$Age == "0-17"] <- "15"
X_train$Age[X_train$Age == "18-25"] <- "21"
X_train$Age[X_train$Age == "26-35"] <- "30"
X_train$Age[X_train$Age == "36-45"] <- "40"
X_train$Age[X_train$Age == "46-50"] <- "48"
X_train$Age[X_train$Age == "51-55"] <- "53"
X_train$Age[X_train$Age == "55+"] <- "60"
X_train$Age <- as.integer(X_train$Age)
```


Converting stay in current city to numeric
```{r}
X_train$Stay_In_Current_City_Years[X_train$Stay_In_Current_City_Years == "4+"] <- "4"
X_train$Stay_In_Current_City_Years <- as.integer(X_train$Stay_In_Current_City_Years)
```

converting Gender to binary
```{r}
X_train$Gender <- ifelse(X_train$Gender == "F", 1, 0)
```


Feature representing the count of each user
```{r}
user_count <- ddply(X_train, .(User_ID), nrow)
names(user_count)[2] <- "User_Count"
X_train <- merge(X_train, user_count, by="User_ID")
```

Feature representing the count of each product
```{r}
product_count <- ddply(X_train, .(Product_ID), nrow)
names(product_count)[2] <- "Product_Count"
X_train <- merge(X_train, product_count, by="Product_ID")
```


Feature representing the average Purchase of each product
```{r}
product_mean <- ddply(X_train, .(Product_ID), summarize, Product_Mean=mean(Purchase))
X_train <- merge(X_train, product_mean, by="Product_ID")

```



feature representing the proportion of times the user purchases the product more than the product's average
```{r}
X_train$flag_high <- ifelse(X_train$Purchase > X_train$Product_Mean,1,0)
user_high <- ddply(X_train, .(User_ID), summarize, User_High=mean(flag_high))
X_train <- merge(X_train, user_high, by="User_ID")
```



Add User_High column to the original dataset also
```{r}
train$flag_high <- ifelse(X_train$Purchase > X_train$Product_Mean,1,0)
user_high <- ddply(train, .(User_ID), summarize, User_High=mean(flag_high))
train <- merge(train, user_high, by="User_ID")
X_train <- merge(X_train, user_high, by="User_ID")
```


Structure of the train dataset
```{r}
str(train)
```

Structure of the X_train dataset which has newly created 
```{r}
str(X_train)
```



Random sample the data set. Just 5% of the data set was take as a sample to explore 
```{r}
library(caTools)
set.seed(3000)
spl = sample.split(train$User_ID, SplitRatio = 0.05)
Pdata=train[spl,]
Pdata$Marital_Status <- ifelse(Pdata$Marital_Status == 1, 'M', 'UM')
```


How is the distribution of Purchases?.
The distributions right skewed and the mean amd meedian for the sample and the population are approximately same.
```{r}
summary(X_train[spl,]$Purchase)
summary(X_train$Purchase)
```


Logtransfermation 
```{r}
library(gridExtra) 
logScale<-qplot(x=log10(Purchase),data=Pdata) # x ais is not scaled by log10
countScale<-ggplot(aes(x=Purchase),data=Pdata)+geom_histogram()+scale_x_log10() # X-axis is scaled by log 10
grid.arrange(logScale,countScale,ncol=2)
```


Look at the gender inbalance in the data set
```{r}
table(Pdata$Gender)   
```


The below frequency poligon says that the proportion of male customer is higher than female in all range of purchases
```{r}
ggplot(aes(x = Purchase, y = ..count../sum(..count..)), data = subset(Pdata, !is.na(Gender))) + 
  geom_freqpoly(aes(color = Gender), binwidth=10) + 
  scale_x_continuous(limits = c(0, 27723), breaks = seq(0, 27723, 50)) + 
  xlab('Purchases') + 
  ylab('Proportion of customers with that Purchase amount')
```


Lets zoom more on the right side , I think some places feamale shoppers overtake males in the case of higher purchases
```{r}
ggplot(aes(x = Purchase, y = ..count../sum(..count..)), data = subset(Pdata, !is.na(Gender))) + 
  geom_freqpoly(aes(color = Gender), binwidth=10) + 
  scale_x_continuous(limits = c(20000, 27723), breaks = seq(0, 1000, 50)) + 
  xlab('Purchases') + 
  ylab('Proportion of customers with that Purchase amount')
```


This is where the feamle proportions overtake teh Male proportions 
```{r}
ggplot(aes(x = Purchase, y = ..count../sum(..count..)), data = subset(Pdata, !is.na(Gender))) + 
  geom_freqpoly(aes(color = Gender), binwidth=10) + 
  scale_x_continuous(limits = c(1500, 2000), breaks = seq(0, 1000, 50)) + 
  xlab('Purchases') + 
  ylab('Proportion of customers with that Purchase amount')
```


How prices are distributed 
The below graph shows that the prices are spread in a multy model distribution 
```{r}
ggplot(train[spl,],aes(x=Purchase))+geom_histogram(binwidth = 1)
ggplot(train,aes(x=Purchase))+geom_density()
ggplot(train,aes(x=Purchase))+geom_dotplot(binwidth = 100)
```


Purchase by gender, statistical exploration. Females distribution is more skewed right
```{r}
by(Pdata$Purchase,Pdata$Gender,summary)   
```
There is a gender imbalance 


Even though there are more outliers in the feamle box plot, their proportion is less compared to male in the case of higher purchases. 
```{r}
qplot(x=Gender, y=Purchase,data=subset(Pdata, !is.na(Gender)),geom='boxplot')
```



Outlier : is 1.5 times IQR distance from the median.
Set the limits to remove the outliers.
So when we see these plots it is clear that the female customers shoppe less than the male
```{r}
qplot(x=Gender, y=Purchase,data=subset(Pdata, !is.na(Gender)),geom='boxplot',ylim=c(0,15750))  
```



Set the limits to remove the outliers
To match the graph details with the above statistics use the coord_cartesian function
```{r}
qplot(x=Gender, y=Purchase,data=subset(Pdata, !is.na(Gender)),geom='boxplot')+
scale_y_continuous(limits=c(0,15750))  

```


This is to summarize the Purchases by Gender numerically.
Coord cartesian will not exclude any observation while summarizing , they will be just left while displaying on the graph, when a ylim is given
```{r}
by(Pdata$Purchase,Pdata$Gender,summary) 
qplot(x=Gender, y=Purchase,data=subset(Pdata, !is.na(Gender)),geom='boxplot')+coord_cartesian(ylim=c(0,15750))
```



product_count is the number of times the product is appearing. 
This means products appearing from 50 to 500 are sold more frequently compared to 1000 to the end.
This means very less proportion ifrequently sold
```{r}
ggplot(X_train,aes(x=Product_Count))+xlab('Number of times a product sold')+geom_density() 
```




Below plot shows the distribution of purchase frequencies and distribution of Occupation frequencies and age frequencies as weoll 
```{r}
library(gridExtra) 
pl1<-ggplot(X_train[spl,],aes(x=Purchase))+geom_density()+ylab('Frequancy')     
pl2<-ggplot(X_train[spl,],aes(x=Occupation))+geom_density()+ylab('Shopper frequency') 
pl3<-ggplot(X_train[spl,],aes(x=Age))+geom_density()+ylab('Frequancy')
grid.arrange(pl1,pl2,pl3,ncol=1)
```



If we take product frequency(Product_count) , the city C has higher frequency 
```{r}
pl1<-ggplot(X_train[spl,],aes(x=Product_Count,fill=City_Category_A))+geom_density()+facet_grid(City_Category_A ~.)+
     scale_x_continuous(limits = c(0,2000))
pl2<-ggplot(X_train[spl,],aes(x=Product_Count,fill=City_Category_B))+geom_density()+facet_grid(City_Category_B ~.)+
     scale_x_continuous(limits = c(0,2000))
pl3<-ggplot(X_train[spl,],aes(x=Product_Count,fill=City_Category_C))+geom_density()+facet_grid(City_Category_C ~.)+
     scale_x_continuous(limits = c(0,2000))
grid.arrange(pl1,pl2,pl3,ncol=1)
```



When we zoomed more on the right, City B has more number of products that are sold less frequently 
```{r}
pl1<-ggplot(X_train[spl,],aes(x=Product_Count,fill=City_Category_A))+geom_density()+facet_grid(City_Category_A ~.)+
  scale_x_continuous(limits = c(1500,1700))
pl2<-ggplot(X_train[spl,],aes(x=Product_Count,fill=City_Category_B))+geom_density()+facet_grid(City_Category_B ~.)+
  scale_x_continuous(limits = c(1500,1700))
pl3<-ggplot(X_train[spl,],aes(x=Product_Count,fill=City_Category_C))+geom_density()+facet_grid(City_Category_C ~.)+
scale_x_continuous(limits = c(1500,1700))
grid.arrange(pl1,pl2,pl3,ncol=1)
```



How Marrital status is effecting the purchase?

Check for imbalance
```{r}
table(Pdata$Marital_Status)
```
Imbalance observed

```{r}
by(Pdata$Purchase,Pdata$Marital_Status,summary)   
```


How about the purchases in the case of married an unmarried?
Unmarried are the frquent shoppers
```{r}
ggplot(aes(x = Purchase, y = ..count../sum(..count..)), data = subset(Pdata, !is.na(Marital_Status))) + 
  geom_freqpoly(aes(color = Marital_Status), binwidth=10) + 
  scale_x_continuous(limits = c(0, 27723), breaks = seq(0, 27723, 50)) + 
  xlab('Purchases') + 
  ylab('Proportion of customers Married and unmarried')
```




